A comparison of model-based and design-based approaches for spatial data.

### 2021-05-21 inst/SpatialDVM_Manuscript/SpatialDVM_Manuscript.Rmd
grts background updates

@@

Spatially balanced sampling algorithms use spatial information to obtain samples spread out in space. Spatially balanced samples are useful because they tend to yield estimators that are more precise than estimators constructed from an sampling algorithm that is not spatially balanced [@stevens2004spatially; @barabesi2011sampling; @grafstrom2013well; @robertson2013bas; @benedetti2017spatiallyreview; @wang2013design]. Many spatially balanced sampling algorithms exist, including the Generalized Random Tessellation Stratified [@stevens2004spatially], the Local Pivotal Method [@grafstrom2012spatially; @grafstrom2018spatially], Spatially Correlated Poisson Sampling [@grafstrom2012spatiallypoisson], Balanced Acceptance Sampling [@robertson2013bas], Within-Sample-Distance [@benedetti2017spatially], and Halton Iterative Partitioning [@robertson2018halton] algorithms. Here we focus on the Generalized Random Tessellation Stratified (GRTS) algorithm, which has several attractive properties that we discuss next.

@@

describe grts

discuss spatial balance

describe estimation and local neighborhood variance

<!-- \citep{StevensOlsen2003VarianceEstimation, StevensOlsen2004GRTS} -->

>>>>

Several comparisons between model-based and design-based for spatial data have been considered, but they tend to compare spatial models to IRS or SIRS (stratified IRS). No comparison has been made between spatial models and spatially balanced design-based approaches.

@@

Spatially balanced sampling algorithms use spatial information to obtain samples spread out in space. Spatially balanced samples are useful because they tend to yield estimators that are more precise than estimators constructed from an sampling algorithm that is not spatially balanced [@stevens2004spatially; @barabesi2011sampling; @grafstrom2013well; @robertson2013bas; @benedetti2017spatiallyreview; @wang2013design]. To quantify spatial balance, @stevens2004spatially proposed statistics based on Voroni polygons. Many spatially balanced sampling algorithms exist, including the Generalized Random Tessellation Stratified [@stevens2004spatially], the Local Pivotal Method [@grafstrom2012spatially; @grafstrom2018spatially], Spatially Correlated Poisson Sampling [@grafstrom2012spatiallypoisson], Balanced Acceptance Sampling [@robertson2013bas], Within-Sample-Distance [@benedetti2017spatially], and Halton Iterative Partitioning [@robertson2018halton] algorithms. Here we focus on the Generalized Random Tessellation Stratified (GRTS) algorithm, which has several attractive properties that we discuss next.

@@

The GRTS algorithm works by mapping two-dimensional space into one-dimensional space while incorporating some elements of randomness. First a square bounding box is placed over the sample frame and divided into four equally sized squares randomly given an level-one address of 0, 1, 2, or 3. Then the inclusion probabilities of the sampling units in each square are summed. If this sum is at least one in any of the cells, a set of four sub-squares are placed within each square. Each set of four sub-squares are randomly given a second address (a level-two address) of 0, 1, 2, 3. Then the inclusion probabilities in each sub-square are summed, and if any of these sums are at least one, a third level of squares are added in the same manner. This hierarchical addressing continues until the inclusion probabilities in the smallest squares are all less than 1. The squares are then mapped to the one-dimensional line in order of their addresses. The length of this line equals $n$, the desired sample size (the sum of the inclusion probabilities). A uniform random variable, denoted by $u_1$ is simulated in $[0, 1]$ and placed on this line. The location of $u_1$ on the line can be linked to a sampling unit, denoted by $s_1$. Then $u_2 \equiv u_1 + 1$, which can be linked to another sampling unit, denoted by $s_2$. Because the GRTS algorithm requires the inclusion probabilities of the smallest sub-cells to be less than one, $s_1$ and $s_2$ must be distinct. This process continues until the $n$ sampling units have been selected as part of the sample.  Further details are provided by [@stevens2004spatially] and [@dumelle2021spsurvey].

The Horvitz-Thompson estimator [@horvitz1952generalization] and its continuous analog [@cordy1993extension] yield unbiased estimates of population totals (and means). If $\tau$ is a population total, then the Horvitz-Thompson estimator of $\tau$, denoted by $\hat{\tau}_{HT}$, is given by
\begin{align}\label{eq:htest}
  \hat{\tau}_{HT} = \sum_{i = 1}^n Y_i \pi_i^{-1},
\end{align}
where $Y_i$ is the observed value of the $i$th sampling unit, and $\pi_i$ is the inclusion probability of the $i$th sampling unit. Estimating the variance of $\hat{\tau}_{HT}$ directly relies on knowing the probability that $s_i$ and $s_j$ are both included in the sample (second-order inclusion probability), which can be unfeasible to compute. Furthermore, this variance estimator does incorporate spatial locations. To address these challenges, [@stevens2003variance] proposed an alternative variance estimator: the local neighborhood variance estimator. The local neighborhood variance estimator does not require second-order inclusion probabilities. It also incorporates spatial locations, which tends to reduce the estimated variance of compared $\hat{\tau}_{HT}$ to a variance estimator ignoring spatial location [@stevens2003variance].

### 2021-05-21 inst/SpatialDVM_Manuscript/SpatialDVM_Manuscript.tex
grts background updates

Spatially balanced sampling algorithms use spatial information to obtain
samples spread out in space. Spatially balanced samples are useful
because they tend to yield estimators that are more precise than
estimators constructed from an sampling algorithm that is not spatially
balanced ((Barabesi and Franceschi, 2011; Benedetti et al., 2017;
Grafström and Lundström, 2013; Robertson et al., 2013; Stevens Jr and
Olsen, 2004; Wang et al., 2013)). Many spatially balanced sampling
algorithms exist, including the Generalized Random Tessellation
Stratified (Stevens Jr and Olsen, 2004), the Local Pivotal Method
(Grafström et al., 2012; Grafström and Matei, 2018), Spatially
Correlated Poisson Sampling (Grafström, 2012), Balanced Acceptance
Sampling (Robertson et al., 2013), Within-Sample-Distance (Benedetti and
Piersimoni, 2017), and Halton Iterative Partitioning (Robertson et al.,
2018) algorithms. Here we focus on the Generalized Random Tessellation
Stratified (GRTS) algorithm, which has several attractive properties
that we discuss next.

@@

accomodates stratification, unequal selection probabilities, oversample
sites It has also recently been updated to accommodate legacy
(historical) sites, minimum distance between sites, and a nearest
neighbor replacement sites comp efficient software

@@

@@

@@

>>>>

Spatially balanced sampling algorithms use spatial information to obtain
samples spread out in space. Spatially balanced samples are useful
because they tend to yield estimators that are more precise than
estimators constructed from an sampling algorithm that is not spatially
balanced (Barabesi and Franceschi, 2011; Benedetti et al., 2017;
Grafström and Lundström, 2013; Robertson et al., 2013; Stevens Jr and
Olsen, 2004; Wang et al., 2013). To quantify spatial balance, Stevens Jr
and Olsen (2004) proposed statistics based on Voroni polygons. Many
spatially balanced sampling algorithms exist, including the Generalized
Random Tessellation Stratified (Stevens Jr and Olsen, 2004), the Local
Pivotal Method (Grafström et al., 2012; Grafström and Matei, 2018),
Spatially Correlated Poisson Sampling (Grafström, 2012), Balanced
Acceptance Sampling (Robertson et al., 2013), Within-Sample-Distance
(Benedetti and Piersimoni, 2017), and Halton Iterative Partitioning
(Robertson et al., 2018) algorithms. Here we focus on the Generalized
Random Tessellation Stratified (GRTS) algorithm, which has several
attractive properties that we discuss next.

@@

The GRTS algorithm works by mapping two-dimensional space into
one-dimensional space while incorporating some elements of randomness.
First a square bounding box is placed over the sample frame and divided
into four equally sized squares randomly given an level-one address of
0, 1, 2, or 3. Then the inclusion probabilities of the sampling units in
each square are summed. If this sum is at least one in any of the cells,
a set of four sub-squares are placed within each square. Each set of
four sub-squares are randomly given a second address (a level-two
address) of 0, 1, 2, 3. Then the inclusion probabilities in each
sub-square are summed, and if any of these sums are at least one, a
third level of squares are added in the same manner. This hierarchical
addressing continues until the inclusion probabilities in the smallest
squares are all less than 1. The squares are then mapped to the
one-dimensional line in order of their addresses. The length of this
line equals \(n\), the desired sample size (the sum of the inclusion
probabilities). A uniform random variable, denoted by \(u_1\) is
simulated in \([0, 1]\) and placed on this line. The location of \(u_1\)
on the line can be linked to a sampling unit, denoted by \(s_1\). Then
\(u_2 \equiv u_1 + 1\), which can be linked to another sampling unit,
denoted by \(s_2\). Because the GRTS algorithm requires the inclusion
probabilities of the smallest sub-cells to be less than one, \(s_1\) and
\(s_2\) must be distinct. This process continues until the \(n\)
sampling units have been selected as part of the sample. Further details
are provided by (Stevens Jr and Olsen, 2004) and (Dumelle et al., 2021).

The Horvitz-Thompson estimator (Horvitz and Thompson, 1952) and its
continuous analog (Cordy, 1993) yield unbiased estimates of population
totals (and means). If \(\tau\) is a population total, then the
Horvitz-Thompson estimator of \(\tau\), denoted by \(\hat{\tau}_{HT}\),
is given by \begin{align}\label{eq:htest}
  \hat{\tau}_{HT} = \sum_{i = 1}^n Y_i \pi_i^{-1},
\end{align} where \(Y_i\) is the observed value of the \(i\)th sampling
unit, and \(\pi_i\) is the inclusion probability of the \(i\)th sampling
unit. Estimating the variance of \(\hat{\tau}_{HT}\) directly relies on
knowing the probability that \(s_i\) and \(s_j\) are both included in
the sample (second-order inclusion probability), which can be unfeasible
to compute. Furthermore, this variance estimator does incorporate
spatial locations. To address these challenges, (Stevens Jr and Olsen,
2003) proposed an alternative variance estimator: the local neighborhood
variance estimator. The local neighborhood variance estimator does not
require second-order inclusion probabilities. It also incorporates
spatial locations, which tends to reduce the estimated variance of
compared \(\hat{\tau}_{HT}\) to a variance estimator ignoring spatial
location (Stevens Jr and Olsen, 2003).

@@

\leavevmode\hypertarget{ref-cordy1993extension}{}%
Cordy, C.B., 1993. An extension of the horvitz---thompson theorem to
point sampling from a continuous universe. Statistics \& Probability
Letters 18, 353--362.

@@

\leavevmode\hypertarget{ref-horvitz1952generalization}{}%
Horvitz, D.G., Thompson, D.J., 1952. A generalization of sampling
without replacement from a finite universe. Journal of the American
statistical Association 47, 663--685.

@@

\leavevmode\hypertarget{ref-stevens2003variance}{}%
Stevens Jr, D.L., Olsen, A.R., 2003. Variance estimation for spatially
balanced samples of environmental resources. Environmetrics 14,
593--610.
